---
title: "R Packages for SDG 6.2 data"
subtitle: "Development and MSc Data Science project"
author: 
    - Linda Karani
    - Lars Schöbitz
format: 
   revealjs:
       smaller: true
       transition: slide
       background-transition: zoom
       css: ["theme/theme.css"]
       theme: moon
       echo: false
---

## Agenda

1. R Package Development
2. Master Thesis Project
    - Fit models for estimates
        - Prepare primary indicators estimates based on JMP estimation rules
        - Prepare secondary indicator service levels
    - Extract r squared and p-values from models
        - Identify those with low values < 0.20 and high values > 0.80
        - Plots exploring the OLS regression using two countries with low and high rsq values
    - Plots showing the model fit coefficients(r.squared and p value)
    - Fit alternative models and assess goodness of fit
    - Next steps
3. Questions for Discussion

```{r, include=FALSE, message=FALSE, warning=FALSE}
#Loading necessary libraries

library(tidyverse)
library(modelr)
library(broom)

```

```{r, include=FALSE, message=FALSE, warning=FALSE }
#Getting data

#jmpraw <- read_rds("data/2020-09-30_jmp_sanitation_raw_data.rds")
jmpraw <- read_csv("https://raw.githubusercontent.com/larnsce/washr/master/data/derived_data/2022-10-18_jmp_sanitation_raw_data.csv")

jmp_measures_of_model_fit <- read_csv(here::here("data/derived/jmp_primary_indicator_estimates_measures_of_model_fit.csv"))

jmp_all_mean_value <-read_csv(here::here("data/derived/jmp_all_mean_estimates.csv"))
```

# R Package Development {background-color="#eee8d5"}

# Master Thesis Project {background-color="#eee8d5"}

## Masters thesis outline

::: incremental

### Objectives

- Generate sanitation estimates for rural and urban from raw packaged data using documented methods
     - only the primary indicators and derived secondary service levels:
         - basic sanitation service
         - limited sanitation service
         - unimproved sanitation service
         - no sanitation service

- Fit different statistical models to assess goodness of fit

### Status

- Managed to derive estimates for the primary indicators and derived secondary
service levels

:::


```{r}
#Filter for R2 less than 0.30
options(scipen=999, digits = 2)


high_rsq <- jmp_measures_of_model_fit %>% 
    group_by(var_short) %>% 
    filter(r.squared > 0.80)

low_rsq <- jmp_measures_of_model_fit %>% 
    group_by(var_short) %>% 
    filter(r.squared < 0.20)

```

# Plot linear fit for different countries


## OLS Regression : r squared \< 0.20

## Afghanistan sewer connection

::: columns
::: {.column width="80%"}

```{r ols-regression, warning=FALSE, echo=FALSE, message=FALSE}


plot1 <- jmp_all_mean_value %>% 
    filter(var_long %in% c("Sewer") & iso3 == "AFG") %>%
    ggplot(aes(x = year, y = mean_value)) + geom_point() + facet_wrap(~residence) +geom_smooth(method = "lm", se = FALSE)+
    labs(x = "year",
         y = "% coverage",
         title = "Afghanistan Sewer connection in rural and urban contexts")+
    theme_gray() +
    theme(title = element_text(size = 12), 
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 12, face = "bold"))

plot1



```
:::

::: {.column width="20%"}

-   Afghanistan r.sq rural = 0.0426018
-   Afghanistan r.sq urban = 0.0170736
:::
:::



## Tanzania open defecation

::: columns
::: {.column width="80%"}

```{r , warning=FALSE, echo=FALSE, message=FALSE}

plot2 <- jmp_all_mean_value %>% 
    filter(var_long %in% c("Open defecation") & iso3 == "TZA") %>%
    ggplot(aes(x = year, y = mean_value)) + geom_point() + facet_wrap(~residence) +geom_smooth(method = "lm", se = FALSE)+
    labs(x = "year",
         y = "% coverage",
         title = "Tanzania Open defecation in rural and urban contexts")+
    theme_gray() +
    theme(title = element_text(size = 12), 
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 12, face = "bold"))

plot2

```

:::

::: {.column width="20%"}

-   Tanzania r.sq rural    = 0.1824534
-   Tanzania r.sq urban    = 0.0296412	
:::
:::



## OLS Regression : r squared \> 0.80

## Argentina Sewer connection OLS regression

::: columns
::: {.column width="80%"}

```{r ols-regression2, warning=FALSE, echo=FALSE, message=FALSE}



plot11 <- jmp_all_mean_value %>% 
    filter(var_long %in% c("Sewer") & iso3 == "ARG") %>%
    ggplot(aes(x = year, y = mean_value)) + geom_point() + facet_wrap(~residence) +geom_smooth(method = "lm", se = FALSE)+
    labs(x = "year",
         y = "% coverage",
         title = "Argentina Sewer connection in rural and urban contexts")+
    theme_gray() +
    theme(title = element_text(size = 12), 
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 12, face = "bold"))

plot11


```
:::

::: {.column width="20%"}

-   Argentina r.sq rural = 0.82
-   Argentina r.sq urban = 0.87
:::
:::




## Brazil Improved sanitation OLS regression

::: columns
::: {.column width="80%"}

```{r, warning=FALSE, echo=FALSE, message=FALSE}

plot21 <- jmp_all_mean_value %>% 
    filter(var_long %in% c("Improved") & iso3 == "BRA") %>%
    ggplot(aes(x = year, y = mean_value)) + geom_point() + facet_wrap(~residence) +geom_smooth(method = "lm", se = FALSE)+
    labs(x = "year",
         y = "% coverage",
         title = "Brazil Improved Sanitation in rural and urban contexts")+
    theme_gray() +
    theme(title = element_text(size = 12), 
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 12, face = "bold"))

plot21

```

:::

::: {.column width="20%"}
-   Brazil r.sq rural = 0.90
-   Brazil r.sq urban = 0.87
:::
:::



# Plotting Model coefficients

## Model fit : R squared {auto-animate="true" auto-animate-easing="ease-in-out"}

```{r, warning=FALSE, echo=FALSE, message=FALSE}


jmp_measures_of_model_fit %>% 
    ggplot(aes(x = var_long, y = r.squared)) + geom_boxplot(fill = "blue") +
    geom_jitter(width = 0.2)+
    facet_wrap(~residence)+
    labs(title = "Model coefficients", x = "Primary indicator",
         subtitle = "r.squared")




```

## Model fit : P value {auto-animate="true" auto-animate-easing="ease-in-out"}

```{r, warning=FALSE, echo=FALSE, message=FALSE}
jmp_measures_of_model_fit %>% 
    ggplot(aes(x = var_long, y = p.value)) + geom_boxplot(fill = "blue") +  
    geom_jitter(width = 0.2)+  
    facet_wrap(~residence)+
    labs(title = "Model coefficients", x = "Primary indicator",
         subtitle = "pvalue")

    
```

# Alternative models

-   Fitting different models to the data to compare goodness of fit

```{r , warning=FALSE, include=FALSE, echo=FALSE, message=FALSE}
# Fitting a polynomial of order 4


poly4model <- jmp_all_mean_value %>% 
    filter(iso3  %in% c("AFG", "TZA")) %>% 
    nest_by(iso3, residence, var_short) %>% 
    mutate(model = list(lm(mean_value ~ poly(year,4, raw = TRUE), data = data)))


rsquared <- poly4model|> 
    summarise(rsq = summary(model)$r.squared) 


  
```

```{r , warning=FALSE, include=FALSE, echo=FALSE, message=FALSE}
#Splines

library(splines)

splinemodel <- jmp_all_mean_value %>% 
    filter(iso3  %in% c("AFG", "TZA")) %>% 
    nest_by(iso3, residence, var_short) %>% 
    mutate(model = list(lm(mean_value ~ bs(year, knots = quantile(jmp_all_mean_value$year, p = c(0.25,0.5,0.75))), data = data)))


rsquared <- splinemodel|> 
    summarise(rsq = summary(model)$r.squared) 

```

```{r , warning=FALSE,include=FALSE, echo=FALSE, message=FALSE}
# Generalized Additive Model

library(mgcv)
library(kableExtra)
library(gratia)



gam_model <- jmp_all_mean_value %>% 
    filter(iso3  %in% c("AFG", "TZA")) %>% 
    group_by(iso3, residence, var_short) %>%
    nest()
    
##Create a function to fit the models    

country_gam <- function(df) {
    gam(mean_value ~ year, family=gaussian, data = df, na.omit=TRUE)
    }
    
##Apply country loess to each element in our df

gam_model <- gam_model %>% 
    mutate(model = map(data, country_gam))

#Add residuals to the model

gam_model <- gam_model %>% 
    mutate(
        resids = map2(data, model, add_residuals))

#unnest

resids <- unnest(gam_model,resids)

#Get statistics using glance

gam_model %>% 
    mutate(glance = map(model,broom::glance)) %>% 
    unnest(glance)


```



```{r , warning=FALSE, include=FALSE,echo=FALSE, message=FALSE, include=FALSE}
#Loess



loess_model <- jmp_all_mean_value %>% 
    filter(iso3  %in% c("AFG", "TZA")) %>% 
    group_by(iso3, residence, var_short) %>%
    nest()
    
##Create a function to fit the models    

country_loess <- function(df) {
    loess(mean_value ~ year, data = df,span = 0.85, degree = 2, normalize = TRUE)
    }
    
##Apply country loess to each element in our df

loess_model <- loess_model %>% 
    mutate(model = map(data, country_loess))

loess_model$model


```



# Plot using a 4th order polynomial

## Afghanistan Sewer connection

::: columns
::: {.column width="80%"}
```{r , warning=FALSE, echo=FALSE, message=FALSE}
plot3 <- jmp_all_mean_value %>% 
    filter(var_long %in% c("Sewer") & iso3 == "AFG") %>%
    ggplot(aes(x = year, y = mean_value)) + geom_point() + facet_wrap(~residence) +
    stat_smooth(method = "lm", formula =  y ~ poly(x,4, raw = TRUE))+
    labs(x = "year",
         y = "% coverage",
         title = "AFG Sewer connection in rural and urban contexts")+
    theme_gray() +
    theme(title = element_text(size = 12), 
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 12, face = "bold"))

plot3
```

:::

::: {.column width="20%"}
-   Afghanistan r.sq rural = 0.043
-   Afghanistan r.sq urban = 0.022

:::
:::



## Tanzania open defecation

::: columns
::: {.column width="80%"}

```{r , warning=FALSE, echo=FALSE, message=FALSE}

plot4 <- jmp_all_mean_value %>% 
    filter(var_long %in% c("Open defecation") & iso3 == "TZA") %>%
    ggplot(aes(x = year, y = mean_value)) + geom_point() + facet_wrap(~residence) +
    stat_smooth(method = "lm", formula =  y ~ poly(x,4, raw = TRUE))+
    labs(x = "year",
         y = "% coverage",
         title = "Tanzania Open Defecation in rural and urban contexts")+
    theme_gray() +
    theme(title = element_text(size = 12), 
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 12, face = "bold"))

plot4

```
:::

::: {.column width="20%"}

-   Tanzania r.sq rural    = 0.487
-   Tanzania r.sq urban    = 0.033
:::
:::

# Plot using splines


## Afghanistan sewer connection

::: columns
::: {.column width="80%"}
```{r, warning=FALSE, echo=FALSE, message=FALSE}
plot5 <- jmp_all_mean_value %>% 
    filter(var_long %in% c("Sewer") & iso3 == "AFG") %>%
    ggplot(aes(x = year, y = mean_value)) + geom_point() + facet_wrap(~residence) +
    stat_smooth(method = "lm", formula =  y ~ splines::bs(x, df = 3))+
    labs(x = "year",
         y = "% coverage",
         title = "AFG Sewer connection in rural and urban contexts")+
    theme_gray() +
    theme(title = element_text(size = 12), 
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 12, face = "bold"))

plot5
```

:::

::: {.column width="20%"}

-   Afghanistan r.sq rural = 1.00
-   Afghanistan r.sq urban = 1.00

:::
:::



## Tanzania Open defecation

::: columns
::: {.column width="80%"}


```{r, warning=FALSE, echo=FALSE, message=FALSE}
plot6 <- jmp_all_mean_value %>% 
    filter(var_long %in% c("Open defecation") & iso3 == "TZA") %>%
    ggplot(aes(x = year, y = mean_value)) + geom_point() + facet_wrap(~residence) +
    stat_smooth(method = "lm", formula =  y ~ splines::bs(x, df = 3))+
    labs(x = "year",
         y = "% coverage",
         title = "Tanzania Open defecation in rural and urban contexts")+
    theme_gray() +
    theme(title = element_text(size = 12), 
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 12, face = "bold"))

plot6

```
:::

::: {.column width="20%"}

-   Tanzania r.sq rural    = 0.65
-   Tanzania r.sq urban    = 0.37
:::
:::

# Plot using GAM

## Afghanistan sewer connection

::: columns
::: {.column width="80%"}

```{r , warning=FALSE, echo=FALSE, message=FALSE}
plot7 <- jmp_all_mean_value %>% 
    filter(var_long %in% c("Sewer") & iso3 == "AFG") %>%
    ggplot(aes(x = year, y = mean_value)) + geom_point() + facet_wrap(~residence) +
    stat_smooth(method = gam, formula =  y ~ s(x, k=3))+
    labs(x = "year",
         y = "% coverage",
         title = "AFG Sewer connection in rural and urban contexts")+
    theme_gray() +
    theme(title = element_text(size = 12), 
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 12, face = "bold"))

plot7
```

:::

::: {.column width="20%"}

-   Afghanistan BIC rural = 18.5
-   Afghanistan BIC urban = 36.1

:::
:::



## Tanzania open defecation

::: columns
::: {.column width="80%"}


```{r , warning=FALSE, echo=FALSE, message=FALSE}
plot8 <- jmp_all_mean_value %>% 
    filter(var_long %in% c("Open defecation") & iso3 == "TZA") %>%
    ggplot(aes(x = year, y = mean_value)) + geom_point() + facet_wrap(~residence) +
    stat_smooth(method = gam, formula =  y ~ s(x, k=3))+
    labs(x = "year",
         y = "% coverage",
         title = "Tanzania open defecation in rural and urban contexts")+
    theme_gray() +
    theme(title = element_text(size = 12), 
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 12, face = "bold"))


plot8
```


:::

::: {.column width="20%"}

-   Tanzania BIC rural = 87.6
-   Tanzania BIC urban = 49.2
:::
:::

# Plotting using LOESS

## Afghanistan sewer connection

::: columns
::: {.column width="80%"}

```{r, warning=FALSE, echo=FALSE, message=FALSE}

plot9 <- jmp_all_mean_value %>% 
    filter(var_long %in% c("Sewer") & iso3 == "AFG") %>%
    ggplot(aes(x = year, y = mean_value)) + geom_point() + facet_wrap(~residence) +
    stat_smooth(method = loess, span = 0.90)+
    labs(x = "year",
         y = "% coverage",
         title = "AFG Sewer connection in rural and urban contexts")+
    theme_gray() +
    theme(title = element_text(size = 12), 
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 12, face = "bold"))

plot9
```

:::

::: {.column width="20%"}

-   Afghanistan RSE rural = 0.043
-   Afghanistan RSE urban = 0.022
:::
:::



## Tanzania open defecation

::: columns
::: {.column width="80%"}

```{r, warning=FALSE, echo=FALSE, message=FALSE}
plot10 <- jmp_all_mean_value %>% 
    filter(var_long %in% c("Open defecation") & iso3 == "TZA") %>%
    ggplot(aes(x = year, y = mean_value)) + geom_point() + facet_wrap(~residence) +
    stat_smooth(method = loess, span = 0.90)+
    labs(x = "year",
         y = "% coverage",
         title = "Tanzania open defecation in rural and urban contexts")+
    theme_gray() +
    theme(title = element_text(size = 12), 
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 12, face = "bold"))

plot10

```
:::

::: {.column width="20%"}

-   Tanzania RSE rural = 0.487
-   Tanzania RSE urban = 0.033
:::
:::
