---
title: "Research workshop summary slides"
format: 
   revealjs:
       transition: slide
       background-transition: fade
       css: ["theme/theme.css"]
       theme: moon
       echo: false
---

# DATA SCIENCE WORKSHOP

# SDG 6.2 Reproducibility

## Outline {.smaller}

Overall vision (raw data, estimates code, jmp estimates package)

Raw data package documentation (proof-of-concept website)

-   Statistics of the raw data

Master Thesis Project

-   Input: Data Package
-   Status
-   What has worked? What hasn't?
-   Plots exploring the statistics
-   Recommendations
-   Gaps
-   Learning
-   Next steps

```{r}
#Loading necessary libraries

library(tidyverse)
library(modelr)
library(broom)

```


```{r}
#Getting data

jmpraw <- read_rds("data/2020-09-30_jmp_sanitation_raw_data.rds")

```

```{r, include=FALSE, message=FALSE, warning=FALSE}

jmp_urban_rural <- jmpraw %>%  
    filter(residence != "national") %>% 
    filter(var_long %in% c("Open defecation", "Improved", "Shared", "Septic" , "Sewer"))

jmp_urban_rural

jmp_all_mean_value <- jmp_urban_rural |> 
    group_by(iso3, residence, var_long, year) |> 
    summarise(mean_value = mean(value)) %>% 
    ungroup()


jmp_all_mean_value

min_max_year_iso3 <- jmp_all_mean_value |> 
    group_by(iso3, var_long) |> 
    summarise(
        min = min(year),
        max = max(year)
    ) 

diff_smaller_five_year_vec <- min_max_year_iso3 |> 
    mutate(diff = max - min) |> 
    filter(diff < 5) |>
    pull(iso3) 


diff_greater_four_year_vec <- min_max_year_iso3 |> 
    mutate(diff = max - min) |> 
    filter(diff >= 5) |>
    pull(iso3)

options(scipen=999, digits = 2)

jmp_models <- jmp_all_mean_value %>% 
    filter(iso3 %in% diff_greater_four_year_vec) |> 
    nest_by(iso3, residence, var_long) %>% 
    mutate(model = list(lm(mean_value ~ year, data = data)))

jmprsquared <- jmp_models |> 
    summarise(rsq = summary(model)$r.squared) 


jmp_models |> 
    summarise(broom::glance(model))

jmp_models |> 
    summarise(broom::tidy(model))





```


## Masters thesis progress

::: {.incremental}

Objectives

- Generate sanitation estimates from raw packaged data using documented methods
- Fit different statistical models to assess goodness of fit

Status

- Managed to derive estimates for the primary indicators

:::

# Reproducibility : Comparing our results to JMP estimates

- Show plots comparing estimates

# Model Fit

## Plotting the distribution of r squared

```{r, warning=FALSE, echo=FALSE, message=FALSE}

#Plotting R squared


jmprsquared %>% 
    ggplot(aes(x = var_long, y = rsq)) + geom_boxplot(fill = "pink") + facet_wrap(~residence)
    

```

```{r, warning=FALSE, echo=FALSE, message=FALSE, include=FALSE}

jmpfilter <- jmprsquared %>% 
    filter(var_long %in% c("Sewer", "Shared", "Improved") & rsq < 0.35)

jmpfilter

```
# Case study : Plot fit of different countries

## Plotting coverage for countries with a low rsquared for var_long equal to sewer/shared

```{r , warning=FALSE, echo=FALSE, message=FALSE}

library(gridExtra)

plot1 <- jmp_all_mean_value %>% 
    filter(var_long %in% c("Sewer") & iso3 == "AFG") %>%
    ggplot(aes(x = year, y = mean_value)) + geom_point() + facet_wrap(~residence) +geom_smooth(method = "lm", se = FALSE)+
    labs(x = "year",
         y = "% coverage",
         title = "AFG Sewer connection in rural and urban contexts")+
    theme_gray() +
    theme(title = element_text(size = 12), 
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 12, face = "bold"))


plot2 <- jmp_all_mean_value %>% 
    filter(var_long %in% c("Improved") & iso3 == "ZWE") %>%
    ggplot(aes(x = year, y = mean_value)) + geom_point() + facet_wrap(~residence) +geom_smooth(method = "lm", se = FALSE)+
    labs(x = "year",
         y = "% coverage",
         title = "Zimbambwe Improved Sanitation in rural and urban contexts")+
    theme_gray() +
    theme(title = element_text(size = 12), 
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 12, face = "bold"))

grid.arrange(plot1, plot2)

```

# Model performance

## Fitting different models to the data to compare performance

```{r , warning=FALSE, echo=FALSE, message=FALSE}
# Fitting a polynomial of order 4


poly4model <- jmp_all_mean_value %>% 
    filter(var_long %in% c("Sewer", "Shared", "Improved")) %>% 
    nest_by(iso3, residence, var_long) %>% 
    mutate(model = list(lm(mean_value ~ poly(year,4, raw = TRUE), data = data)))


rsquared <- poly4model|> 
    summarise(rsq = summary(model)$r.squared) 


  
```

```{r , warning=FALSE, echo=FALSE, message=FALSE}
#Splines

library(splines)

splinemodel <- jmp_all_mean_value %>% 
    filter(var_long %in% c("Sewer", "Shared", "Improved")) %>% 
    nest_by(iso3, residence, var_long) %>% 
    mutate(model = list(lm(mean_value ~ bs(year, knots = quantile(jmp_all_mean_value$year, p = c(0.25,0.5,0.75))), data = data)))


rsquared <- splinemodel|> 
    summarise(rsq = summary(model)$r.squared) 

```

```{r , warning=FALSE, echo=FALSE, message=FALSE}
# Generalized Additive Model

library(mgcv)

GAMmodel <- jmp_all_mean_value %>% 
    filter(var_long %in% c("Sewer", "Shared", "Improved")) %>% 
    mutate(model = list(gam(mean_value ~ s(year) , data = .)))


    
```

```{r , warning=FALSE, echo=FALSE, message=FALSE, include=FALSE}
#Loess

loessmodel <- loess(mean_value ~ year, data = jmp_all_mean_value, span = 0.5, degree = 2)



predict(loessmodel)

```

# Plots

## Plot using a 4th order polynomial

```{r , warning=FALSE, echo=FALSE, message=FALSE}

plot3 <- jmp_all_mean_value %>% 
    filter(var_long %in% c("Sewer") & iso3 == "AFG") %>%
    ggplot(aes(x = year, y = mean_value)) + geom_point() + facet_wrap(~residence) +
    stat_smooth(method = "lm", formula =  y ~ poly(x,4, raw = TRUE))+
    labs(x = "year",
         y = "% coverage",
         title = "AFG Sewer connection in rural and urban contexts")+
    theme_gray() +
    theme(title = element_text(size = 12), 
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 12, face = "bold"))

plot4 <- jmp_all_mean_value %>% 
    filter(var_long %in% c("Improved") & iso3 == "ZWE") %>%
    ggplot(aes(x = year, y = mean_value)) + geom_point() + facet_wrap(~residence) +
    stat_smooth(method = "lm", formula =  y ~ poly(x,4, raw = TRUE))+
    labs(x = "year",
         y = "% coverage",
         title = "Zimbambwe Improved Sanitation in rural and urban contexts")+
    theme_gray() +
    theme(title = element_text(size = 12), 
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 12, face = "bold"))

grid.arrange(plot3,plot4)

```

## Plot using splines

```{r, warning=FALSE, echo=FALSE, message=FALSE}

plot5 <- jmp_all_mean_value %>% 
    filter(var_long %in% c("Sewer") & iso3 == "AFG") %>%
    ggplot(aes(x = year, y = mean_value)) + geom_point() + facet_wrap(~residence) +
    stat_smooth(method = "lm", formula =  y ~ splines::bs(x, df = 3))+
    labs(x = "year",
         y = "% coverage",
         title = "AFG Sewer connection in rural and urban contexts")+
    theme_gray() +
    theme(title = element_text(size = 12), 
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 12, face = "bold"))

plot6 <- jmp_all_mean_value %>% 
    filter(var_long %in% c("Improved") & iso3 == "ZWE") %>%
    ggplot(aes(x = year, y = mean_value)) + geom_point() + facet_wrap(~residence) +
    stat_smooth(method = "lm", formula =  y ~ splines::bs(x, df = 3))+
    labs(x = "year",
         y = "% coverage",
         title = "Zimbambwe Improved Sanitation in rural and urban contexts")+
    theme_gray() +
    theme(title = element_text(size = 12), 
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 12, face = "bold"))

grid.arrange(plot5,plot6)


```

## Plot using GAM

```{r , warning=FALSE, echo=FALSE, message=FALSE}

plot7 <- jmp_all_mean_value %>% 
    filter(var_long %in% c("Sewer") & iso3 == "AFG") %>%
    ggplot(aes(x = year, y = mean_value)) + geom_point() + facet_wrap(~residence) +
    stat_smooth(method = gam, formula =  y ~ s(x, k=3))+
    labs(x = "year",
         y = "% coverage",
         title = "AFG Sewer connection in rural and urban contexts")+
    theme_gray() +
    theme(title = element_text(size = 12), 
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 12, face = "bold"))


plot8 <- jmp_all_mean_value %>% 
    filter(var_long %in% c("Improved") & iso3 == "ZWE") %>%
    ggplot(aes(x = year, y = mean_value)) + geom_point() + facet_wrap(~residence) +
    stat_smooth(method = gam, formula =  y ~ s(x, k=3))+
    labs(x = "year",
         y = "% coverage",
         title = "Zimbambwe Improved Sanitation in rural and urban contexts")+
    theme_gray() +
    theme(title = element_text(size = 12), 
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 12, face = "bold"))


grid.arrange(plot7,plot8)
```

## Plotting using LOESS

```{r, warning=FALSE, echo=FALSE, message=FALSE}

plot9 <- jmp_all_mean_value %>% 
    filter(var_long %in% c("Sewer") & iso3 == "AFG") %>%
    ggplot(aes(x = year, y = mean_value)) + geom_point() + facet_wrap(~residence) +
    stat_smooth(method = loess, span = 0.90)+
    labs(x = "year",
         y = "% coverage",
         title = "AFG Sewer connection in rural and urban contexts")+
    theme_gray() +
    theme(title = element_text(size = 12), 
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 12, face = "bold"))

plot10 <- jmp_all_mean_value %>% 
    filter(var_long %in% c("Improved") & iso3 == "ZWE") %>%
    ggplot(aes(x = year, y = mean_value)) + geom_point() + facet_wrap(~residence) +
    stat_smooth(method = loess, span = 0.90)+
    labs(x = "year",
         y = "% coverage",
         title = "Zimbambwe Improved Sanitation in rural and urban contexts")+
    theme_gray() +
    theme(title = element_text(size = 12), 
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 12, face = "bold"))

grid.arrange(plot9,plot10)

```
