---
title: "Thesis report"
format: html
editor: source
bibliography: msc-thesis-lkarani.json
csl: apa.csl
---

```{r}
library(tidyverse)
```

# Introduction

## Background

The Sustainable Development Goals (SDGs), otherwise known as the Global Goals, are a set of objectives within a universal agreement to end poverty, protect all that makes the planet habitable, and ensure that all people enjoy peace and prosperity, now and in the future [@morton2017sustainable].The Sustainable development Goal 6 is one of the 17 Sustainable Development Goals that addresses access to clean water and sanitation. Access to water, sanitation and hygiene is a basic right that is recognized as playing a vital role in the 2030 Agenda by improving progress in other areas such as health, education and poverty reduction(UNEP,n.d). The  international monitoring of drinking water and sanitation has been in effect since the 1930s when the League of Nations Health Organization was responsible for monitoring [@bartram2014global]. The World Health Organization(WHO) then took over and now The Joint Monitoring Programme (JMP) which is a body that consists of The World Health Organization(WHO) and The United Nations Children‘s Fund (UNICEF) is responsible for global monitoring of water, sanitation, and hygiene estimates since 1990 [@supply2015progress]. They produce a total of 26 indicators for 232 countries related to water, sanitation and hygiene at the national, regional and global level(JMP,2017). The indicators include the percentage of households using different technologies or levels of service which is further disaggregated by rural and urban settings [@wolf2013exploration].

## Calculation of estimates

To compute indicator estimates for water, sanitation and hygiene(WASH), the JMP relies on data from a variety of national data sources including: national censuses, national household surveys and administrative data. Data is manually extracted from these sources and organized into a standardized format using individual spreadsheets (the JMP country files). These country files contain this data in a structured format, which allows scripted programming languages to use the data and compute indicator estimates using statistical methods. For each of the 232 countries, the estimates are generated by fitting an ordinary least squares regression line to a series of data points [@perez-foguet2017compositional]. 

STATA is used as a tool to read the raw data from the country file, fit the model to compute the estimates, and write the results back into the county file. The methods are described in a report published in March 2018 (Joint Monitoring Programme, 2018). The JMP country files allow for a transparent review of raw data sources and published estimates. The STATA script that does the data transformation remains unpublished and closed to the public. However, there is great potential to publish the code that produces the estimates as open source software to facilitate others to verify the results, build upon them, or contribute to further development of statistical methods. This brings about the concept of Reproducibility. 

## Reproducibility



## Calculation of estimates
JMPs mission is “to be the trusted source of global, regional and national data on sustainable access to safe drinking water and basic sanitation for use by governments, donors, international organizations and civil society”. Producing accurate global estimates requires accurate data to effectively monitor progress on Sustainable Development Goals(SDGs) and enhance quality evidence - based decision making.

# Problem statement



# Results & Discussion



```{r}
#| child: "sections/03-02-results-model-fit.qmd"
```






# Conclusions

# Recommendations

# References
