---
title: "Thesis report"
format: html
editor: source
bibliography: msc-thesis-lkarani.json
csl: apa.csl
---

```{r}
library(tidyverse)
```

# Introduction

## Background

The Sustainable Development Goals (SDGs), otherwise known as the Global Goals, are a set of objectives within a universal agreement to end poverty, protect all that makes the planet habitable, and ensure that all people enjoy peace and prosperity, now and in the future [@morton2017sustainable].The Sustainable development Goal 6 is one of the 17 Sustainable Development Goals that addresses access to clean water and sanitation. Access to water, sanitation and hygiene is a basic right that is recognized as playing a vital role in the 2030 Agenda by improving progress in other areas such as health, education and poverty reduction(UNEP,n.d). 2 billion people worldwide still lack access to safe drinking water and 3.6 billion lack access to safe sanitation [@ProgressHouseholdDrinking]. In order to achieve the 2030 target of universal coverage, current rates will need to be significantly increased. Monitoring of SDG 6 estimates is an essential part of tracking progress and informing development by identifying areas which need improvement.
The  international monitoring of drinking water and sanitation has been in effect since the 1930s when the League of Nations Health Organization was responsible for monitoring [@bartram2014global]. The World Health Organization(WHO) then took over and now The Joint Monitoring Programme (JMP) which is a body that consists of The World Health Organization(WHO) and The United Nations Children‘s Fund (UNICEF) is responsible for global monitoring of water, sanitation, and hygiene estimates since 1990 [@supply2015progress]. They produce a total of 26 indicators for 232 countries related to water, sanitation and hygiene at the national, regional and global level(JMP,2017). The indicators include the percentage of households using different technologies or levels of service which is further disaggregated by rural and urban settings [@wolf2013exploration]. 

## Calculation of estimates

To compute indicator estimates for water, sanitation and hygiene(WASH), the JMP relies on data from a variety of national data sources including: national censuses, national household surveys and administrative data. Data is manually extracted from these sources and organized into a standardized format using individual spreadsheets (the JMP country files). These country files contain this data in a structured format, which allows scripted programming languages to use the data and compute indicator estimates using statistical methods. For each of the 232 countries, the estimates are generated by fitting an ordinary least squares regression line to a series of data points [@perez-foguet2017compositional]. 

STATA is used as a tool to read the raw data from the country file, fit the model to compute the estimates, and write the results back into the county file. The methods are described in a report published in March 2018 (Joint Monitoring Programme, 2018). The JMP country files allow for a transparent review of raw data sources and published estimates. The STATA script that does the data transformation remains unpublished and closed to the public. However, there is great potential to publish the code that produces the estimates as open source software to facilitate others to verify the results, build upon them, or contribute to further development of statistical methods. This brings about the concept of Reproducibility. 

## Reproducibility

The term Reproducibility was first noted by Claerbout and Karrenbach [@Claerbout.Karrenbach1992] and has been used in scientific research ever since. According to the National Academies of Sciences, Engineering, and Medicine [@NationalAcademiesofSciences2019], Reproducibility means obtaining consistent results using the same input data, computational steps, methods, and conditions of analysis. Victoria Stodden [@zotero-2005] classifies reproducibility into empirical, computational and statistical. Empirical reproducibility  includes non-computational empirical scientific experiments and observations. Computational reproducibility on the other hand entails detailed information on data, code and processes. Lastly, Statistical reproducibility considers new measures to assess the reliability and stability of statistical inferences, including developing new validation measures and expanding the field of uncertainty quantification to develop measures of statistical confidence and a better understanding of sources of error, especially when large multi-source datasets or massive simulations are involved [@Community.Scriberia2022] . Reproducibility has gradually been gaining traction in the past decade with attempts to reproduce research in the social science field slowly increasing. When research is made transparent, other researchers are able to easily use the same processes, data and code to verify old findings or build upon it.

Producing accurate global estimates requires statistically sound and reproducible methods to effectively monitor progress on Sustainable Development Goals(SDGs) and enhance quality evidence - based decision making [@wolf2013exploration]. This is in line with the Joint Monitoring programme's mission “to be the trusted source of global, regional and national data on sustainable access to safe drinking water and basic sanitation for use by governments, donors, international organizations and civil society” (JMP, n,d). With 2030 fast approaching, it is important that we review the current methods in an effort to strengthen monitoring and reporting. Without reliable data systems, it is difficult to improve water, sanitation and hygiene services or measure progress towards targets. The objective of this study will be to reproduce the JMP sanitation estimates using the JMP's documented methods, review the methodology used to reproduce the estimates and in so doing suggest alternative statistical models that are able to capture the different trajectories present in the data.

# Problem statement

Advancement in SDG 6 depends a lot on a strong foundation of data credibility. Accurate data is the lifeblood of decision-making and the raw material for ensuring accountability [@zotero-2009]. To calculate its estimates, The Joint Monitoring Programme relies on country-level open government data. . This data is stored in multiple formats, which are often not machine-readable (e.g. PDFs), unstructured, and without metadata. Processing this data is time consuming, prone to errors, and difficult to reproduce.This could be overcome by making the data readily accessible in a form that follows the FAIR data principles [@Wilkinson.etal2016].It is therefore crucial that the data shared through open government sources is machine readable to facilitate comparability, more accurate and efficient analysis. In addition, the current JMP method has received criticism as the generated estimates differ from the national country estimates. This is likely due to the underlying data sources or the fact that the estimates are derived from a linear regression which fails to capture different country non linear trends[@bartram2014global]. The JMP uses linear regression due to its simplicity, the limited number data points available and the ease of explanation[@perez-foguet2017compositional]. [@ezbakhe2019estimating ; @fuller2016tracking] propose alternative modeling methods to better monitor the SD6 2030 target.

To overcome the challenges described above, this study will develop a reproducible R package that derives the estimates from the raw data in addition to evaluating alternative statistical model and building a decision tree that map outs appropriate models depending on the number of data points and the trend a country has.

# Research objectives

This study seeks to answer the following questions:

1. Is it feasible to reproduce indicator estimates for SDG indicator 6.2.1 using published raw data and documented methods? 
2. Are alternative statistical models able to accurately assess different country trends and offer better performance to effectively monitor progress of achieving universal water and sanitation(WASH) coverage?

In addition to answering the questions above, this research aims to contribute to better reporting of the SDG 6 estimates by:

1. Opening up the data transformation process of computing JMP’s global estimates by using the same input data in a bid to attain the same results which is in line with its mission of producing accurate global estimates.	
2. Addressing common issues around data management such as data stored in non machine-readable formats, such as PDFs or data being shared by email, and how this may deter progress towards universal access to sanitation (SDG 6.2)
3. Providing a user-friendly R package that follows FAIR principles for data management and stewardship [@Wilkinson.etal2016]  for users to conduct their own analyses from the JMP data to answer new questions and better understand trends in access to sanitation and safe management
4. Developing a decision tree that selects models based on different country data scenarios 

# Literature review

Data makes science possible. Sharing data improves visibility, and makes the research process transparent. This increases trust in the work, and allows for independent reproduction of results. Good data management is not a goal in itself, but rather is the key conduit leading to knowledge discovery and innovation, and to subsequent data and knowledge integration and reuse by the community after the data publication process [@Wilkinson.etal2016]. This section of the study will be two fold : The first section will focus on some of the data management principles as highlighted by different authors that overcome the barrier to reproducibility while the second section with focus on different methodologies employed by researchers to the Water and Sanitation(WASH) raw data to assess trends and monitor progress.

## Principles of data mangement and reproducibility.



# Results & Discussion



```{r}
#| child: "sections/03-02-results-model-fit.qmd"
```






# Conclusions

# Recommendations

# References
