---
title: "JMP raw sanitation data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{my_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This vignette provides an explanation of the variables in the jmp raw data sanitation dataset.

## Variable descriptions

- source : Represents the code associated with the source of the data i.e census, report or survey.
- type : Type of the data i.e census, survey or report
- year : Represents the year when the estimate was recorded
- var_short : Short variable sanitation description
- value : Represents the percentage pf the population using a particular sanitation service
- iso3 : ISO3 country code
- var_long : Represents the sanitation service
- residence : Rural or urban residence
- san_service_chain : Represents the service ladder associated with the sanitation indicator

## Installation

Use the below code to install the package
```{r setup}
remotes::install_github("KaraniLinda/SDG-6-2-Reproducibility/JMPrawdatapackage")

library(JMPrawdatapackage)
```
